\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-07 lang-design.tex
stuff
late binding

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLC_fix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.

\noindentparagraph{Extensibility of inductive types.}

@@

\noindentparagraph{Exhaustivity of induction.}

@@

With inductive types made extensible, exhaustivity of induction becomes
in question, however.
To reconcile the tension without requiring redefinition or rechecking of
case handlers (challenge \ref{chg:extensible-v-exhaustive}),
our design introduces the \lsti{FRecursion} and \lsti{FInduction} commands.
The key idea is to allow case handlers to be added retroactively, should
inductive types be extended, and to allow recursion and induction (which are defined in terms of case handlers) to be late bound.

@@

Like all references to the fields of a family, the meaning of "subst" is late bound.
For example, the type of "st_beta" (modeling the $\beta$-reduction rule)
in family "STLC" refers to "subst".
When "st_beta" is inherited into family "STLC_fix", it has a type that now
refers to the "subst" function in "STLC_fix", where "subst" is defined
by handling all the five cases known to that family.

The \lsti{FInduction} command is similar to \lsti{FRecursion} but
allows cases to be defined in proof mode.
For example, "weakenlem" is proven by induction on the typing relation
"hasty".
Its \lsti{motive} clause suggests that the lemma reads as
\lsti{\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T}.
Upon entering case "ht_abs", for instance, Coq enters proof mode with the goal

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.

\noindentparagraph{Equality on late bound definitions.}

@@

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded: the definition of \lsti{subst}, as a recursive function,
varies across families, yet the proof of the \lsti{ht_unit} case should be
reusable by a derived family regardless of how that family may add
new cases to the definition of \lsti{subst}.
How can the programmer change the goal to \lsti{hasty G tm_unit ty_unit}, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), can still be proved.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Hence, our plugin provides a tactic \lsti{fsimpl}
with which the programmer can rewrite \lsti{subst tm_unit x t'} to
\lsti{tm_unit}.

>>>>

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLC_fix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.
For example, executing the last line, \lsti{Check STLC_fix.typesafe}, will
display the type-safety theorem of the fixpoints extension.

\noindentparagraph{Extending inductive types.}

@@

\noindentparagraph{Ensuring exhaustivity of induction.}

@@

With inductive types made extensible, exhaustivity of induction becomes
in jeopardy, however.
To reconcile the tension without requiring redefinition or rechecking of
case handlers (challenge \ref{chg:extensible-v-exhaustive}),
our design introduces the \lsti{FRecursion} and \lsti{FInduction} commands.
The key idea is to allow case handlers to be added retroactively, should
inductive types be extended, and to allow recursion and induction (which are defined in terms of case handlers) to be late bound.

@@

The \lsti{FInduction} command is similar to \lsti{FRecursion} but
allows cases to be defined in proof mode.
Consider "weakenlem" as an example.
It is proven by induction on the typing relation "hasty".
Its \lsti{motive} clause suggests that the lemma reads as

\begin{lstlisting}[basicstyle=\fontsize{8}{9}\ttfamily]
\forall G t T, hasty G t T -> \forall G', includedin G G' -> hasty G' t T.
\end{lstlisting}

\noindent
Upon entering case "ht_abs", for instance, Coq enters proof mode with the goal

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLC_fix" adds
a constructor to "hasty", the programmer is required to handle this extra case
in the proof of "substlem" in family "STLC_fix".

\noindentparagraph{Late binding of nested names.}

OO inheritance allows the late binding of method names.
Family polymorphism generalizes the power of OO inheritance by allowing the
late binding of all names nested within families, including those referring to types.
We have seen that late binding of "tm" allows the "tm" constructors in "STLC"
to be reused in "STLC_fix" to construct terms containing "tm_fix".

As another example, consider the type of "st_beta" in family "STLC".
It refers to "subst", whose meaning is late bound.
When "st_beta" is inherited into family "STLC_fix", "st_beta" has a type that now
refers to the "subst" function in "STLC_fix", where "subst" is defined
by handling all the five cases known to that family.
Thus, late binding of "subst" allows the derived family to reuse "st_beta"
to $\beta$-reduce abstractions whose body is a fixpoint.

Importantly, a name is late bound only within a family that defines or extends it.
Outside such families, the name can be accessed only by explicitly
specifying a family that contains it.
For example, the last line of \cref{fig:stlc-mechanized} accesses "typesafe"
with "STLC_fix.typesafe".
Executing the command prints

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_fix.typesafe: \forall t t' T, STLC_fix.steps t t' -> STLC_fix.hasty STLC_fix.empty t T ->
                   STLC_fix.value t' \/ \exists t'', STLC_fix.step t' t''.
\end{lstlisting}

\noindent
In this type, all references to nested names are qualified by "STLC_fix", as desired.

\noindentparagraph{Equality on late bound names.}

@@

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded (challenge~\ref{chg:latebind-v-defeq}):
the definition of \lsti{subst}, as a recursive function,
varies across families, yet a derived family should be able to reuse the
proof of the \lsti{ht_unit} case
even when it has to modify the definition of \lsti{subst}.
Without the ability to unfold \lsti{subst},
how can the programmer make progress in this proof, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), can still be proved.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Based on this insight, our plugin automatically generates and proves
propositional equalities for recursive functions defined with \lsti{FRecursion}.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the generated
propositional equalities.

Note that definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound.
In contrast, outside those families, \lsti{subst} can be referenced only
by specifying a family that contains it,
such as \lsti{STLC.subst} or \lsti{STLC_fix.subst}.
So as far as Coq's type checker is concerned, \lsti{tm_unit} and
\lsti{STLC_fix.subst tm_unit x t'} are the same thing---Coq equates them
definitionally by unfolding \lsti{STLC_fix.subst} and performing
$\beta$-reduction.

\noindentparagraph{Overriding.}

\noindentparagraph{Ensuring consistency.}

Example of a circular proof

\noindentparagraph{Composing families.}

\noindentparagraph{Tactics for eliminating inductive types.}

\noindentparagraph{Limitations.}
